Finite element analysis of multi-dimensional and simplified models for beams

### 2024-04-16 final_seminar/main.tex
final

@@

        \begin{frame}{Detailed Comparison: Timoshenko Beam Model vs. Two-Dimensional Model}

            \textbf{Objective of Comparison:}
            \begin{itemize}
              \item To validate the efficacy of the Timoshenko beam theory against a more comprehensive two-dimensional model, which accounts for plane stress conditions and detailed stress distribution across the beam's cross-section.
            \end{itemize}

            \textbf{Methodology:}
            \begin{itemize}
              \item Utilized finite element methods to calculate natural frequencies and modes, providing a direct comparison of results from both models.
              \item Both models analyzed under the same boundary conditions and loading scenarios to ensure consistency in comparison.
            \end{itemize}

            \textbf{Key Findings:}
            \begin{itemize}
              \item \textbf{Close Agreement}: Timoshenko model results closely match those from the two-dimensional model, with most eigenvalues showing less than 3\% deviation.
              \item \textbf{Model Reliability}: Demonstrates high reliability of the Timoshenko model for predicting beam behavior under various loading conditions, particularly for lower vibration modes.
            \end{itemize}

@@

        \begin{frame}
            \textbf{Model Advantages:}
            \begin{itemize}
              \item \textbf{Timoshenko Model}: Offers significant improvements in accuracy over simpler models like Euler-Bernoulli, especially in scenarios where beam lengths are relatively short and shear deformations cannot be ignored.
              \item \textbf{Two-Dimensional Model}: Provides a more rigorous analysis by accounting for two-dimensional stress effects, which are important for comprehensive structural analyses.
            \end{itemize}

            \textbf{Practical Implications:}
            \begin{itemize}
              \item \textbf{Engineering Applications}: Timoshenko model is particularly advantageous for practical engineering applications, providing a good balance between computational efficiency and accuracy.
              \item \textbf{Advanced Analyses}: For critical engineering tasks where detailed stress distribution is crucial, the two-dimensional model serves as a more suitable analytical tool.
            \end{itemize}

            \textbf{Conclusions and Recommendations:}
            \begin{itemize}
              \item \textbf{Preferred Use of Timoshenko Model}: Recommended for most practical engineering applications involving beams, where it strikes an optimal balance between accuracy and simplicity.
              \item \textbf{When to Use Two-Dimensional Model}: Advisable for in-depth structural analyses where precise stress distribution is required, such as in complex assemblies or where materials with varying properties are involved.
              \item This comparison serves as an intermediate step towards the more comprehensive analysis involving the comparison of two-dimensional and three-dimensional models.
            \end{itemize}

    \subsection{Empirical comparison of beams}
        \begin{frame}{Literature study - Empirical comparison of beams}
            Consider the article \textit{On the valid frequency range of Timoshenko beam theory} [\cite{SP06}]\footfullcite{SP06}
        \end{frame}

        \begin{frame}{Setup and Testing Method}
            \textbf{Objective:}
            \begin{itemize}
                \item To validate the frequency predictions of Timoshenko beam theory (TBT) for different end conditions and frequencies.
            \end{itemize}

@@

            \textbf{Key Setup Features:}
            \begin{itemize}
                \item The experimental beam was suspended and excited by two carbon-fibre loops attached to piezoelectric transducers.
                \item Resonant frequencies detected by sweeping the frequency range with a network analyzer.
            \end{itemize}
        \end{frame}

        \begin{frame}{Numerical Results and Discussion}
            \textbf{Results Summary:}
            \begin{itemize}
                \item Excellent agreement between measured, ANSYS, and RUS frequencies below the cut-off frequency.
                \item Above the cut-off frequency, TBT predictions deviate increasingly from measured and simulated frequencies.
            \end{itemize}

            \textbf{Discussion:}
            \begin{itemize}
                \item For free-free end conditions, discrepancies arise above the cut-off frequency, suggesting limitations in the TBT's applicability.
                \item TBT predictions were increasingly inaccurate for higher modes (n >= 8), exceeding the typical engineering tolerance.
            \end{itemize}

            \textbf{Conclusion:}
            \begin{itemize}
                \item The study suggests a cut-off frequency as a reasonable upper bound for TBT predictions.
                \item This research serves as an intermediate step, setting the stage for further comparison between two-dimensional and three-dimensional models.
            \end{itemize}
        \end{frame}

@@

            \only<2>{
            Dividing both sides by $T(t)$ gives
            \begin{eqnarray*}
                b(x,v) = - \frac{T''(t)}{T(t)}c(x,v) \implies  \frac{T''(t)}{T(t)} \ \ \ \textrm{must be constant.}
            \end{eqnarray*}}

@@

            ***************Verduideliking********************

            Solutions for the Eigenvalue Problem:
            \begin{eqnarray*}
                u = \sum_{i=1}^{\infty} a_i x_i = \sum_{i=1}^{\infty} c(u, x_i)x_i.
            \end{eqnarray*}

            Solutions for the Ordinary Differential Equation:
            \begin{flalign}
                T_n(t) &=  A_n \cos(\sqrt{\lambda_n} t) + B_n \sin(\sqrt{\lambda_n} t) \quad \textrm{ if } \lambda_n > 0, \label{lambda_1}
            \end{flalign}

            Formal series solution for the general vibration problem:
            \begin{eqnarray}
                u(t) = \sum_{n=1}^{\infty} T_n(t)x_n.
            \end{eqnarray}
        \end{frame}

\section{Numerical Results}
    \subsection{Finite Element Analysis}
        \begin{frame}{Eigenvalue problem - Two and Three-dimensional models}
            General eigenvalue problem for the two- and three-dimensional models:

@@

        \begin{frame}{Eigenvalue problem - Timoshenko model}
            General eigenvalue problem for the Timoshenko beam model:

>>>>

            \textbf{Objective of Comparison:}
            \begin{itemize}
              \item To validate the efficacy of the Timoshenko beam theory against a more realistic two-dimensional model.
            \end{itemize}

            \textbf{Methodology:}
            \begin{itemize}
              \item Use modal analysis and the Finite Element Method (FEM) to calculate natural frequencies and modes, providing a direct comparison of results from both models.
            \end{itemize}

            \textbf{Key Findings:}
            \begin{itemize}
              \item Timoshenko model results closely match those from the two-dimensional model, with most eigenvalues showing less than 3\% deviation, depending on the shape of the beam.
              \item The shape of the beams significantly influences the accuracy of the Timoshenko model.
            \end{itemize}

@@

        \begin{frame}{Literature study - Comparison of linear beam theories}

            \textbf{Conclusions and Recommendations:}
            \begin{itemize}
              \item For practical applications, the Timoshenko beam model is a suitable approximation for two-dimensional model.
              \item This article excluded damping.
              \item This comparison serves as an intermediate step towards the more comprehensive analysis involving the comparison of Timoshenko beam model and a three-dimensional model. A direct comparison is preferred but this will introduce additional complexities.
            \end{itemize}

            \textbf{In this presentation:}
            \begin{itemize}
              \item We will replicate the results of this article using the Finite Element Method (FEM) and modal analysis.
              \item We will improve the results by increasing the accuracy to 5 significant digits.
              \item We extent the article to also include the comparison of the two-dimensional model with a three-dimensional model.
            \end{itemize}

@@

    \subsection{Empirical comparison of beams}
        \begin{frame}{Literature study - Empirical comparison of beams}
            Consider the article \textit{On the valid frequency range of Timoshenko beam theory} [\cite{SP06}]

            \textbf{Objective:}
            \begin{itemize}
                \item To validate the frequency predictions of Timoshenko beam theory for different end conditions and frequencies.
            \end{itemize}

@@

            \footfullcite{SP06}
        \end{frame}

        \begin{frame}{Setup and Testing Method}

            \textbf{Key Setup Features:}
            \begin{itemize}
                \item The experimental beam was suspended and excited by two carbon-fibre loops attached to piezoelectric transducers.
                \item Resonant frequencies detected by sweeping the frequency range with a network analyzer.
            \end{itemize}

            \begin{figure}[h!]
                \centering
                \begin{tikzpicture}[scale=0.8]

                    \draw[line width = 0.25mm] (2,0) -- (7,2);
                    \draw[line width = 0.25mm] (2,1) -- (7,3);
                    \draw[line width = 0.25mm] (1,1) -- (6,3);

                    \draw[line width = 0.25mm] (2,0) -- (2,1);
                    \draw[line width = 0.25mm] (1,0) -- (1,1);
                    \draw[line width = 0.25mm] (2,0) -- (1,0);
                    \draw[line width = 0.25mm] (2,1) -- (1,1);

                    \draw[line width = 0.25mm] (6,3) -- (7,3);
                    \draw[line width = 0.25mm] (7,3) -- (7,2);

                    \draw(2.1,0) parabola (2.5,4);
                    \draw(1,0.1) parabola (0.6,4);

                    \draw(6.9,1.9) parabola (7.4,4.9);
                    \draw[line width = 0.15mm] (5.7,2.9) -- (5.5,4.9);

                \end{tikzpicture}
            \end{figure} 
        \end{frame}

        \begin{frame}{Numerical Results and Discussion}
            \textbf{Results Summary:}
            \begin{itemize}
                \item Excellent agreement between measured, ANSYS, and RUS frequencies below the cut-off frequency.
            \end{itemize}

            \textbf{Some results from the article:}
            \begin{table}[htbp]
                \makebox[\textwidth]{
                \begin{tabular}{||c|c|cc|cc|c||}
                    \hline
                    {n} & {Measured} & {3D FEM} & {\% Error} & {Timoshenko Beam} & {\% Error} & Side \\
                    \hline
                    1     & 27359.6 & 27417.6 & 0,21\% & 27407.1 & 0,17\% & Flexible \\
                            & 27423.8 & 27515.6 & 0,33\% & 27505.3 & 0,30\% & Stiff \\
                    2     & 60862   & 60882.0 & 0,03\% & 60851.1 & -0,02\% & Flexible \\
                           & 61098.3 & 61022.0 & -0,12\% & 60992.1 & -0,17\% & Stiff \\
                    3     & 97609.5 & 97734.4 & 0,13\% & 97796.0 & 0,19\% & Flexible \\
                          & 97852.4 & 97881.5 & 0,03\% & 97945.4 & 0,09\% & Stiff \\
                    4     & 161494 & 131658 & 0,12\% & 132277 & 0,60\% & Flexible \\
                          & 131732 & 131675 & -0,04\% & 132308 & 0,44\% & Stiff \\
                    5     & 161352 & 161390 & 0,02\% & 163547 & 1,36\% & Stiff \\
                          & 161538 & 161517 & -0,01\% & 163611 & 1,31\% & Flexible \\
                    6     & 165183 & 164887 & -0,18\% & 169108 & 2,38\% & Stiff \\
                          & 165598 & 165398 & -0,12\% & 169634 & 2,44\% & Flexible \\
                    7     & 194863 & 194933 & 0,04\% & 202352 & 3,84\% & Flexible \\
                             & 194973 & 195032 & 0,03\% & 202115 & 3,66\% & Stiff \\
                    8     & 195869 & 195977 & 0,06\% & 203319 & 3,80\% & Stiff \\
                          & 195908 & 196097 & 0,10\% & 203518 & 3,88\% & Flexible \\
                    9     & 213501 &       &       & 241202 & 12,97\% & Flexible \\
                          & 213635 &       &       & 241067 & 12,84\% & Stiff \\
              {10 and 11} & 220556 &       &       & 247954 & 12,42\% & Flexible \\
                          & 220702 &       &       & 281542 & 27,57\% & Flexible \\
                          & 221010 &       &       & 247782 & 12,11\% & Stiff \\
                          & 221092 &       &       & 281408 & 27,28\% & Stiff \\
                    \hline
                \end{tabular}%
            }
            \end{table}%
        \end{frame}

@@

            \only<2->{
            Dividing both sides by $T(t)$ gives
            \begin{eqnarray*}
                b(x,v) = - \frac{T''(t)}{T(t)}c(x,v) \implies  \frac{T''(t)}{T(t)} \ \ \ \textrm{must be constant.}
            \end{eqnarray*}}

@@

            Using these assumptions, the authors prove that there exists a complete orthonormal sequence of eigenvectors for the eigenvalue problem with a corresponding sequence of eigenvalues. These eigenvalues are positive and the orthogonality is with respect to the bilinear form $c$. Also the sequence of normalized eigenvectors ${x_i}$ forms an orthonormal basis in $W$ and sequence of eigenvalues ${\lambda_i}$ is an infinite sequence with $\lambda_n \rightarrow \infty$ as $n \rightarrow \infty$.

            Following these results, for any $u \in V$, $\displaystyle u = \sum_{i=1}^{\infty} a_i x_i$. These coefficients $a_i$ are generalized Fourier coefficients of $u$ with respect to the eigenvectors $x_i$. Therefore, for any $u \in V$,
            \begin{eqnarray*}
                u = \sum_{i=1}^{\infty} a_i x_i = \sum_{i=1}^{\infty} c(u, x_i)x_i.
            \end{eqnarray*}

            Now that the eigenvalue problem has many solutions, the following ordinary differentiable equation can be considered,
            \begin{eqnarray*}
                T_n'' + \lambda T_n = 0. 
            \end{eqnarray*}
        \end{frame}

        \begin{frame}
            Since this differential equation is a simple second order differential equation, $T_n(t)$ has the following possible solutions:
            \begin{flalign}
                T_n(t) &=  A_n \cos(\sqrt{\lambda_n} t) + B_n \sin(\sqrt{\lambda_n} t) \quad \textrm{ if } \lambda_n > 0, \label{lambda_1}
            \end{flalign}

            Then combining the solutions of the eigenvalue problem and the differential equation, the formal series solution for the boundary value problem is
            \begin{eqnarray}
                u(t) = \sum_{n=1}^{\infty} T_n(t)x_n. \label{eq:1D_Model:ModalAnalysisSeriesSolution}
            \end{eqnarray}

\section{Numerical Results}
    \subsection{Finite Element Analysis}
        \begin{frame}{Eigenvalue problem - Timoshenko model}
            General eigenvalue problem for the Timoshenko beam model:

        \begin{frame}{Eigenvalue problem - Two and Three-dimensional models}
            General eigenvalue problem for the two- and three-dimensional models:

@@
